A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-11-09 main.tex
Small tweaks - avoid merge conflicts
Formatting for genetic epi submission; using endfloat

To describe and compare PC-lasso and LMM-lasso, we consider an $n \times p$ matrix of genotype data $\bG$ for $n$ subjects and $p$ SNPs, where each $G_{ij} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. In order to equitably compare SNP effects, regardless of allele frequency, in practice $\bG$ is standardized. We denote the standardized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 \pi_j) / \sqrt{\pi_j (1 - \pi_j)}$ and $\pi_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, $\bX_0$ represents an $n \times p_0$ matrix of unpenalized covariates and the intercept. Thus, at a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include observed covariates such as age, sex, or smoking history, which are included in the model without penalty. We assume that the $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of these covariates, additive SNP effects, environmental effects, and observational noise, and consider the linear model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

Although we derive these quantities under model \eqref{eqn:lmm_reg}, where $\bZ = \bX$, these results also hold for the more general model \eqref{eqn:model}, where $\bZ \ne \bX$. We see this by decomposing $\bZ \bgamma$ into components in and orthogonal to the column space of the projection matrix of $\bX$.  Let $\boldsymbol{P}_{\bX} = \bX (\bX^\T \bX)^{-1} \bX^\T$ denote the projection matrix of $\bX$, with $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denoting the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then there exist $\boldsymbol{\tau}, \boldsymbol{\psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$, where $\boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$ and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$.  This in turn means that
\begin{align}
    \label{eqn:partition_e}    \by &= \bX\bbeta + \bZ\bgamma + \beps \notag \\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps \notag\\
    &=  \bX \bbeta{^*} + \beps{^*};
\end{align}
note that $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat^*$ and $\boldsymbol{\psi}$ is absorbed by the residual error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.  This result provides the explicit mathematical justification for our statement from Section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure.

@@

We begin by comparing the ordinary lasso, PC-lasso, and LMM-lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-lasso, and LMM-lasso was evaluated based on mean squared error (MSE) \anna{MSPE sims running - maybe add here}. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 SNR setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-lasso perform comparably. PC-lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

@@

In this work, we have reviewed the concept of population structure in a framework which \anna{formally differentiates between effects of population structure and those of the environment}. We have provided a broad outline of existing methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two popular approaches for correcting for population structure in genetic data, \anna{now applied within a novel penalized framework}: PC-lasso and LMM-lasso.

>>>>

To describe and compare PC-lasso and LMM-lasso, we consider an $n \times p$ matrix of genotype data $\bG$ for $n$ subjects and $p$ SNPs, where each $G_{ij} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. In order to equitably compare SNP effects, regardless of allele frequency, in practice $\bG$ is standardized. We denote the standardized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 \pi_j) / \sqrt{2\pi_j (1 - \pi_j)}$ and $\pi_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, $\bX_0$ represents an $n \times p_0$ matrix of unpenalized covariates and the intercept. Thus, at a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include observed covariates such as age, sex, or smoking history, which are included in the model without penalty. We assume that the $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of these covariates, additive SNP effects, environmental effects, and observational noise, and consider the linear model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

Although we derive these quantities under model \eqref{eqn:lmm_reg}, where $\bZ = \bX$, these results also hold for the more general model \eqref{eqn:model}, where $\bZ \ne \bX$. We see this by decomposing $\bZ \bgamma$ into components in and orthogonal to the column space of the projection matrix of $\bX$.  Let $\boldsymbol{P}_{\bX} = \bX (\bX^\T \bX)^{-1} \bX^\T$ denote the projection matrix of $\bX$, with $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denoting the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then there exist $\boldsymbol{\tau}, \boldsymbol{\psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$, where $\bX \boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$ and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$.  This in turn means that
\begin{align}
    \label{eqn:partition_e}    \by &= \bX\bbeta + \bZ\bgamma + \beps \notag \\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps \notag\\
    &=  \bX \bbeta{^*} + \beps{^*};
\end{align}
note that $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat^*$ and $\boldsymbol{\psi}$ is absorbed by the residual error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.  This result provides the explicit mathematical justification for our statement from Section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure.

@@

We begin by comparing the ordinary lasso, PC-lasso, and LMM-lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-lasso, and LMM-lasso was evaluated based on mean squared error (MSE). Figure \ref{fig:mse} presents these results for all three data types in a 1:1 SNR setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-lasso perform comparably. PC-lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

@@

In this work, we have reviewed the concept of population structure in a framework which formally differentiates between effects of population structure and those of the environment. We have provided a broad outline of existing methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two popular approaches for correcting for population structure in genetic data, now applied within a novel penalized framework: PC-lasso and LMM-lasso.
